---
title: "RNA-seq Analysis for GSE267325"
author: "B282025"
date: "25/03/2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Analysis of GSE267325 RNA-seq data using R 

First, we load the required libraries and files for the pipeline.
```{r init, echo=FALSE}
#Load the required libraries & load the files
library(Rsubread)
library(edgeR)
library(limma)
library(gplots)
library(DESeq2)
library(affy)
library(QuasR)
library(pheatmap)
library(EnhancedVolcano)
library(RColorBrewer)
load("mytable_feaures")

#We now have files for the tutorial and also data loaded into the workspace
ls()
dir()
```

Next, we will use R to generate the targets file.
```{r input_processing, echo=FALSE}
# Some further processing to access count data and column names.
# Reference: FGT ICA Example Analysis
table_rnaseq <- (mytable_feaures)$counts # this is the count data
colnames_rnaseq <- colnames(table_rnaseq)
print(colnames_rnaseq)

# Generate an 'adf data frame' for the annotation targets file.  
# Reference: FGT ICA Example Analysis
adf <-data.frame(Colnames = colnames_rnaseq)
library(stringr)
locc <- str_locate_all(pattern="Aligned.sortedByCoord.out.bam",adf$Colnames)
adf$GEOAccession <- substr(adf$Colnames,1,locc[[1]][,"start"]-1)

# Sample Annotation from the GEO page for this study: 
# https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE267325

# GSM8263915  dLAM1-CSF
# GSM8263916  dLAM2-CSF
# GSM8263917  dLAM3-CSF
# GSM8263918  Mono-MΦ1-CSF
# GSM8263919  Mono-MΦ2-CSF
# GSM8263920  Mono-MΦ3-CSF

#My samples are in the same order as the annotation, so I will add the short names to the annotation table
# Reference: https://stackoverflow.com/questions/5293715/how-to-use-greek-symbols-in-ggplot2
adf$ShortNames <- c("dLAM1","dLAM2","dLAM3","Mono_M\u03A61","Mono_M\u03A62","Mono_M\u03A63")

# Create a factor column for the macrophage type
cell_type <- c("dLAM", "dLAM", "dLAM", "Mono_M\u03A6", "Mono_M\u03A6", "Mono_M\u03A6")
adf$CellType <- factor(cell_type)

# Add more annotation data to the adf data frame...
# BioSample IDs
adf$BioSample <- c("SAMN41382413", "SAMN41382412", "SAMN41382411", "SAMN41382410", "SAMN41382409", "SAMN41382408")

# Experiment IDs
adf$Experiment <- c("SRX24727943","	SRX24727944", "SRX24727945", "SRX24727946", "SRX24727947", "SRX24727948")

#Add colours to the annotation table
# Reference: FGT ICA Example Analysis
# Reference: https://stackoverflow.com/questions/42288075/how-to-autogenerate-color-vector-from-a-numeric-vector
# Reference: https://r-graph-gallery.com/38-rcolorbrewers-palettes.html
z <- factor(adf$CellType)
adf$Colours <- (brewer.pal(length(levels(z)),"Accent"))[z]

# Confirm column names of the study match the rownames of the annotation.
# Reference: FCT ICA Example Analysis
rownames(adf) <-adf$Colnames
if(identical(colnames_rnaseq,rownames(adf))==FALSE){
  print("Names do not match")
}else{
  print("Names match.. OK to proceed")
}

# Switch to shortnames to help with the figures
# Reference: FGT ICA Example Analysis 
#We can only swap this way because column order to consistent with adf
rownames(adf) <-adf$ShortNames
colnames(table_rnaseq) <-adf$ShortNames
colnames_rnaseq <-colnames(table_rnaseq)

# Save the annotated data frame as a targets file
# Reference: Code adapted from Using R for Data Science Lecture 2
write.csv(adf, file="targets_GSE267325.csv", row.names=TRUE)

```
Once we have loaded the data and generated the annotation (saved as targets file) we are ready to run some further analysis.

## Loading the Data into DESeq2
DESeq2 will used for the statistical analysis and the normalisation of the data. 
```{r loading_deseq, echo=FALSE}
# Reference for code chunk: FGT ICA Example Analysis

# Create object for DESeq2
dds_rnaseq <- DESeqDataSetFromMatrix(countData = table_rnaseq, colData = adf, design = ~ CellType)

# Check the object dimensions
dim(dds_rnaseq)
head(rownames(dds_rnaseq))
colnames(dds_rnaseq)

#Check the object type
class(dds_rnaseq)
typeof(dds_rnaseq)

#Check what this object contains
slotNames(dds_rnaseq)
colnames(colData(dds_rnaseq))
colnames(assay(dds_rnaseq))

```

#Quality Filtering Genes
Apply filters to remove genes that do not pass quality filters. In this case, we filter for genes with counts greater than ten.
```{r apply_filters, echo=FALSE}
# First filter based upon counts
# Reference for code chunk: FGT ICA Example Analysis
keep <- rowSums(counts(dds_rnaseq)) > 10
dds_rnaseq <- dds_rnaseq[keep,]
nrow(dds_rnaseq)
```

## Normalisation of Data
```{r normalisation, echo=TRUE}
# Reference for code chunk: FGT ICA Example Analysis
# Size factor estimation for normalisation- these are used to scale by library size when data is exported.
dds_rnaseq <- estimateSizeFactors(dds_rnaseq)
sizeFactors(dds_rnaseq)

```

DESeq2 also provides methods that outputs data with various normalisations. Here, we generate multiple different normalisations and compare between thenm.

```{r plot_normalised, echo=FALSE}
# Reference for code chunk: FGT ICA Example Analysis

#Normalised log counts
counts_rnaseq <-log2(counts(dds_rnaseq, normalized=TRUE)+1)

#fpm is similar to cpm and discussed here https://rdrr.io/bioc/DESeq2/man/fpm.html.  Here the data is log transformed
fpm_rnaseq <-log(fpm(dds_rnaseq)+1) #this is also normalised by the sizeFactors

#rlog ad vsd are discussed here https://genomebiology.biomedcentral.com/articles/10.1186/s13059-014-0550-8
vsd_rnaseq <- vst(dds_rnaseq, blind = T)
#head(vsd_rnaseq, 3)

rld_rnaseq <- rlog(dds_rnaseq, blind = T)
#head(rld_rnaseq, 3)
```

##Plots with Normalised Data.
```{r build_plots, echo=FALSE}
# Reference for code chunk: FGT ICA Example Analysis
#Boxplot for non-normalized data
counts_un <-counts(dds_rnaseq, normalized=FALSE)+1
boxplot(log2(counts_un), main="Non-Normalized Count Data", ylab="log2 Expression", col=adf$Colours, las = 2)

counts_un_rlog<-rlog(counts_un)
boxplot(counts_un_rlog, main="Non-Normalized Count Data", ylab="rlog Expression", col=adf$Colours, las = 2)

# Boxplots for normalized data
na.rm=T
boxplot(counts_rnaseq, main="Normalized Count Data", ylab="log2 Counts", col=adf$Colours, las = 2)
boxplot(fpm_rnaseq, main="FPM Normalized Count Data", ylab="log Counts", col=adf$Colours, las = 2)
boxplot(assay(vsd_rnaseq), main=" VST Normalized Count Data", ylab="log Counts", col=adf$Colours, las = 2)
boxplot(assay(rld_rnaseq), main="rlog Normalized Count Data", ylab="rlog Counts", col=adf$Colours, las = 2)

```

The non-normalized plots do not present any concerns. The samples and replicates groups all have similar means and interquartile ranges.

The different normalization techniques serve different purposes. The rlog normalized data has the most consistent mean and interquartile range spread across samples and it does the best job at shrinking the extreme values. For the rest of the pipeline, we will use rlog.

```{r save_boxplot echo=FALSE}
# Save the rlog normalized boxplot as png for the report
png("normalized_boxplot.png", res=300, width = 2000, height = 1200)
boxplot(assay(rld_rnaseq), main="rlog Normalized Count Data", ylab="rlog Counts", col=adf$Colours, las = 2, cex.axis = 0.7)
dev.off()
```

#Exploratory Data Plots
First, we will view the MvA plots to compare all sample pairs.
```{r mva_plots, echo=FALSE}
# Reference for code chunk: FGT ICA Example Analysis

# Non-normalized rlog counts
mva.pairs(counts_un_rlog, main="MvA Plot for Non-Normalized rlog Count Data", cex=1)

# Normalized rlog counts
mva.pairs(assay(rld_rnaseq), main="MvA Plot for Normalized rlog Count Data", cex=1)

# Save the normalized MvA plot as png for the report
png("MVA_plot.png", res=300, width = 2000, height = 1200)
mva.pairs(assay(rld_rnaseq), main="MvA Plot for Normalized Count Data", cex=1)
dev.off()
```
After normalization, the majority of MvA plots are denser and flatter with points centered around zero, indicating that the normalization worked adequately. 

The authors of the orginal paper used a 2-dimensional PCA. I will generate a 2d plot for comparison, as well as a 3d plot.
```{r pca_explore, echo=FALSE}

# Reference for code chunk: FGT ICA Example Analysis
library(ggplot2)

# 2-dimensional PCA plot
# Reference: https://ggplot2.tidyverse.org/reference/geom_text.html#ref-examples
# Reference: https://www.statology.org/remove-legend-ggplot2/ 
# Reference: https://felixfan.github.io/ggplot2-remove-grid-background-margin/
# Reference: https://ggplot2.tidyverse.org/reference/aes_colour_fill_alpha.html#ref-examples 
ress <-plotPCA(rld_rnaseq, intgroup = c("CellType"))
ress <- ress + 
  geom_text(aes(label = adf$ShortNames),check_overlap=FALSE, size=3, nudge_x=-5, nudge_y=0.5) +
  theme_bw() + 
  labs(title = "PCA Plot of RNA-seq") + 
  theme(aspect.ratio=1/2) +
  theme(legend.position="none")
plot(ress)

```

```{r 3d_pca, echo=FALSE}
library(scatterplot3d)
# Reference for code chunk: FGT ICA Example Analysis

# Perform PCA
pca <- prcomp(t(na.omit(assay(rld_rnaseq))), scale=T)

# Plot the PCA results for the first three principle components
s3d<-scatterplot3d(pca$x[,1:3], pch=19, color=adf$Colours, main = "PCA of RNA-seq")
s3d.coords <- s3d$xyz.convert(pca$x[,1:3])
text(s3d.coords$x, s3d.coords$y, labels = colnames(rld_rnaseq), pos = 3,offset = 0.5, cex=0.5)

# Save PCA plot for report
png("pca_plot_3d.png", res=300, width = 2000, height = 1200)
s3d<-scatterplot3d(pca$x[,1:3], pch=19, color=adf$Colours, main = "PCA of RNA-seq")
s3d.coords <- s3d$xyz.convert(pca$x[,1:3])
text(s3d.coords$x, s3d.coords$y, labels = colnames(rld_rnaseq), pos = 3,offset = 0.5, cex=0.5)
dev.off()

```


## Generate heatmaps with the data

```{r more_heatma-s, echo=FALSE}
# Reference for code chunk: FGT ICA Example Analysis
# Reference: https://r-graph-gallery.com/38-rcolorbrewers-palettes.html 
library("pheatmap")

# Generate a distance matrix
sampleDists <- dist(t(assay(rld_rnaseq)))
#sampleDists

sampleDistMatrix <- as.matrix(sampleDists)
pheatmap(sampleDistMatrix, main="Heatmap of Gene Expression Differences by Sample", color = brewer.pal(11, "PRGn"), display_numbers = FALSE)

# Save as png for report
pheatmap(sampleDistMatrix, main="Heatmap of Gene Expression Differences by Sample", color = brewer.pal(11, "PRGn"), display_numbers = FALSE, file="sample_differences_heatmap.png")
```
The sample distances represent the differences between gene expression between every pair of samples. We see no difference between one cell type and itself, which makes sense. We see that the cell types cluster together. There is more between group variance than within group variance. The dLAM1 sample and the Mono-Mφ3 sample are the most different. This agrees with what we see on the PCA plot. The cell types cluster together.  

## Differential Gene Expression
Here we perform a two factor analysis using DESeq2.
```{r differential_expression, echo=FALSE}
# Reference for code chunk: Modified from FGT ICA Example Analysis
dds_rnaseq <- DESeq(dds_rnaseq)

#We should apply this here https://support.bioconductor.org/p/77461/#107693
#But there is a bug in this version of DESeq2!!
#dds_rnaseq <- lfcShrink(dds_rnaseq, coef=2)

# Specify the contrast
# This gives us a table of differentially expressed genes
result_celltype <-results(dds_rnaseq, contrast=c("CellType", "dLAM", "Mono_M\u03A6"))
write.csv(result_celltype, "results.csv") # save the results to a csv file
results_gsea <- result_celltype # use this copy of the table later for GSEA

# padj is the p-value with a multiple testing correction
# Plot of differential expression
plotMA(result_celltype, main="DESeq2 Dispersion Plot", ylim=c(-10,10))

# Create a table of the differentially expressed genes with padj values < 0.01
table(result_celltype$padj < 0.01)

# Create a table of the differentially expressed genes with padj values < 0.05
table(result_celltype$padj < 0.05)

# Create a table of significant genes
result_celltype_selected <- subset(result_celltype, padj < 0.05)

# Order the significant (padj < 0.05) genes by fold change
result_celltype_selected <-result_celltype_selected[order(abs(result_celltype_selected$log2FoldChange),decreasing = TRUE), ]
head(result_celltype_selected)
```
We find 3657 genes with significant differential expression at a p-adjusted threshold value of 0.01.

We find 5233 genes with significant differential expression at a p-adjusted threshold value of 0.05.

## Make heatmaps from each of the output tables- selected top 10, top 50, and all using fold change
```{r differential_expression_heatmaps, echo=FALSE}
# Reference for code chunk: Modified from FGT ICA Example Analysis

# Make a heatmap for the top 10 differentially expressed genes
top10 <- rownames(result_celltype_selected)[1:10]
# Here we are indexing the rlog normalized count data using the rownames of the top 10 significantly differentially expressed genes
pheatmap(assay(rld_rnaseq)[top10,],
         scale="row",
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         show_rownames=T,
         color = brewer.pal(11, "PRGn"), 
         main="Top 10 Differentially Expressed Genes")

# Make a heatmap for the top 50 differentially expressed genes
top50 <-rownames(result_celltype_selected)[1:50]
# Here we are indexing the rlog normalized count data using the rownames of the top 50 significantly differentially expressed genes
pheatmap(assay(rld_rnaseq)[top50,],
         scale="row",show_rownames=F, 
         color = brewer.pal(11, "PRGn"), 
         main="Top 50 Differentially Expressed Genes")

# Make a heatmap for all of the significant genes
pheatmap(assay(rld_rnaseq)[rownames(result_celltype_selected),],
         scale="row",show_rownames=F,
         color = brewer.pal(11, "PRGn"),
         main="Heatmap of all Significantly Differentially Expressed Genes")

# Save heatmap of all genes
png("heatmap_all.png", res=300, width = 3000, height = 1200)
pheatmap(assay(rld_rnaseq)[rownames(result_celltype_selected),],
         scale="row",show_rownames=F,
         color = brewer.pal(11, "PRGn"),
         main="Heatmap of all Significantly Differentially Expressed Genes")
dev.off()

```

We are comparing resident macrophages to recruited macrophages. These are very different and so it makes sense that like all of the genes are differentially expressed.


## Build Gene Annotation. A serialised is generated in case Ensembl is unavailable

```{r build_gene_annotation, echo=TRUE}
# Reference for code chunk: FGT ICA Example Analysis

#Download Ensembl annotation using BiomaRt and rename the samples
library(biomaRt)
if(file.exists("resultAnnot.RData")){
  load("resultAnnot.RData")
  print("Loaded existing Ensembl annotation...")
}else{
  
  #UK ensembl is being updated  we use a USA mirror, "useast.ensembl.org"
  ensembl_host <-"https://useast.ensembl.org"
  head(biomaRt::listMarts(host = ensembl_host), 15)
  head(biomaRt::listAttributes(biomaRt::useDataset(dataset = "mmusculus_gene_ensembl",mart = useMart("ENSEMBL_MART_ENSEMBL",host = ensembl_host))), 40)
  mart <- biomaRt::useDataset(dataset = "mmusculus_gene_ensembl",mart    = useMart("ENSEMBL_MART_ENSEMBL",host    = ensembl_host))
  resultAnnot <- biomaRt::getBM(values=rownames(dds_rnaseq),attributes =     c("ensembl_gene_id","external_gene_name","chromosome_name","start_position","end_position","description","strand"),filters="ensembl_gene_id",mart=mart)
  save(resultAnnot,file="resultAnnot.RData")
}

```

Merge Annotation with Input Data
```{r merge_gene_annotation, echo=FALSE}
# Reference for code chunk: Modified from FGT ICA Example Analysis

#merge with input data
names <- resultAnnot[,1]
resultAnnot <- as.data.frame(resultAnnot)
rownames(resultAnnot) = names
idx<-match(rownames(dds_rnaseq),rownames(resultAnnot))


# Make sure annotation is in same order
all(rownames(dds_rnaseq) == rownames(resultAnnot))
grr<-resultAnnot[match(rownames(dds_rnaseq), resultAnnot$ensembl_gene_id),]
all(rownames(dds_rnaseq) == rownames(grr))
resultAnnot <-grr
all(rownames(dds_rnaseq) == rownames(resultAnnot))

# The boolean statements above are returning FALSE values so I will check 
# whether most of the genes are matching
# Reference: https://stackoverflow.com/questions/74111575/difference-between-in-and-in-operator-in-r 

# Genes that have a match in the annotation table
sum(rownames(dds_rnaseq) %in% resultAnnot$ensembl_gene_id)

# Genes that don't have a match in the annotation table
sum(!rownames(dds_rnaseq) %in% resultAnnot$ensembl_gene_id)

# It looks like only a few genes (13) are missing annotations


# Make the nice names
nice_names<-paste(resultAnnot$ensembl_gene_id,resultAnnot$external_gene_name, sep = '_')
resultAnnot$nice_names <-nice_names
head(resultAnnot)
all(rownames(dds_rnaseq) == rownames(resultAnnot))

#check names
rld_rnaseq <- rlog(dds_rnaseq, blind = TRUE)#
idx2 <-match(rownames(result_celltype_selected)[1:10],rownames(dds_rnaseq))
plotme <-(rld_rnaseq)[rownames(result_celltype_selected)[1:10],]
rownames(plotme)<-resultAnnot$nice_names[idx2]

# Make heatmap with candidate genes
pheatmap(assay(plotme),
         scale="row",
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         show_rownames=T,
         color = brewer.pal(11, "PRGn"), 
         main="Top 10 Differentially Expressed Genes",
         fontsize_row = 10,
         cellheight =12, cellwidth =12,
         treeheight_row = 30, 
         treeheight_col = 30)

# Save heatmap to a file
png("top10_heatmap.png", res=300, width = 3000, height = 1200)
pheatmap(assay(plotme),
         scale="row",
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         show_rownames=T,
         color = brewer.pal(11, "PRGn"), 
         main="Top 10 Differentially Expressed Genes",
         fontsize_row = 10,
         cellheight =12, 
         cellwidth =12,
         treeheight_row = 20, 
         treeheight_col = 20)
dev.off()

```
## Final Results Table
Since we have annotation that matches gene names to candidates we can also annotate the results table from differential expression.

13 genes have missing annotations as we discovered above. To deal with this, I labeled the missing annotations as "unannotated", rather than removing the genes.

```{r merge_gene_annotation2, echo=FALSE}
# Reference for code chunk: Modified from FGT ICA Example Analysis
#this has the results ordered

#before 
head(result_celltype,50)

idx2 <-match(rownames(result_celltype),rownames(resultAnnot))

# Reference: ChatGPT was consulted to generate some ideas for dealing with missing annotations
names_no_na <- resultAnnot$nice_names[idx2]
names_no_na[is.na(names_no_na)] <- paste0(rownames(result_celltype)[is.na(names_no_na)], "_Unannotated")

rownames(result_celltype) <- names_no_na


#after
print(result_celltype[1:50,])

#load into table viewer
#View(as.data.frame(result_celltype[1:50,]))

```

## The source paper showed candidates in a volcano plot- this is to compare

Current row names are not useful here- we need to use genes but then will discard any
without known names. This is OK here because we are exploring the annotation -genes without
a gene name and so are no use for annotating the data.

We can still keep unannotated genes as candidates!

```{r make_volcano, echo=FALSE}
# Reference for code chunk: Modified from FGT ICA Example Analysis
result_celltype$gene<-resultAnnot$external_gene_name[idx2]

# Reference: Using R for Data Science Tutorial 7b Logical Operators
has_annot <- result_celltype$gene=="" | is.na(result_celltype$gene)
has_annot <-!has_annot
result_celltype<-result_celltype[has_annot,]
rownames(result_celltype)<-result_celltype$gene

#Now we have gene names so we can plot the remaining genes in a volcano plot

# These were highlighted as interesting in the paper
paper_genes <-c("Cd74", "Mrc1", "Cd163", "Dab2", "Lyve1", "Ms4a7", "Cxcl9", "Mmp14", "Apoe", "Arg1", "Cx3cr1", "Il10")

# Confirm they are measured in this annotation
if(sum(paper_genes %in% rownames(result_celltype))==length(paper_genes)){
  print("all genes found in data")
} else {
  print(sum(paper_genes %in% rownames(result_celltype)))
}

#annotated_paper_genes <- paper_genes[paper_genes %in% rownames(result_celltype)]
#print(annotated_paper_genes)
#you'd just include a subset to label if some genes were missing- here they are all present


```

```{r volcano}
# Reference: https://bioconductor.org/packages/devel/bioc/vignettes/EnhancedVolcano/inst/doc/EnhancedVolcano.html 

keyvals.colour <- ifelse(
    result_celltype$log2FoldChange < -1 & result_celltype$padj < 0.05, '#BEAED4',
      ifelse(result_celltype$log2FoldChange > 1 & result_celltype$padj < 0.05, '#7FC97F', 'grey'))
  keyvals.colour[is.na(keyvals.colour)] <- 'grey'
  names(keyvals.colour)[keyvals.colour == '#7FC97F'] <- 'Up'
  names(keyvals.colour)[keyvals.colour == 'grey'] <- 'NS'
  names(keyvals.colour)[keyvals.colour == '#BEAED4'] <- 'Down'

volcano_test <-EnhancedVolcano(result_celltype,
    lab = rownames(result_celltype),
    x = 'log2FoldChange',
    y = 'padj',
    selectLab = paper_genes,
    xlab = bquote(~Log[2]~ 'fold change'),
    pCutoff = 5e-02,
    FCcutoff = 1.0,
    xlim = c(-9, 9),
    ylim = c(0, 90),
    pointSize = 3,
    labSize = 5,
    title = "Differential Expression Results",
    subtitle = "dLAMs vs Mono_M\u03A6",
    labCol = 'black',
    boxedLabels = FALSE,
    colCustom = keyvals.colour,
    colAlpha = 4/5,
    legendPosition = 'top',
    legendLabSize = 15,
    arrowheads = FALSE,
    gridlines.major = FALSE,
    gridlines.minor = FALSE,
    legendIconSize = 5,
    drawConnectors = TRUE,
    widthConnectors = 1,
    colConnectors = 'black')

plot(volcano_test,xlim=800,ylim=800)
png("volcano_test.png",width = 2000, height = 2000, res=300)
plot(volcano_test)
dev.off()

#and look at the table
#View(as.data.frame(result_celltype[annotated_paper_genes,]))
mmp_result <- result_celltype[paper_genes,]
print(as.data.frame(result_celltype[paper_genes,]))

```


Heatmap of the paper genes
```{r Mmp_heatmap, echo=FALSE}
# Reference for code chunk: Modified from FGT ICA Example Analysis
rld_rnaseq <- rlog(dds_rnaseq, blind = TRUE)
paper_ids <- resultAnnot$ensembl_gene_id[resultAnnot$external_gene_name %in% paper_genes]

plotme2 <- (rld_rnaseq)[rownames(rld_rnaseq) %in% paper_ids,]

idx3 <- match(rownames(plotme2), resultAnnot$ensembl_gene_id)
rownames(plotme2) <- resultAnnot$external_gene_name[idx3]


# Make heatmap with paper genes
pheatmap(assay(plotme2),
         scale="row",
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         show_rownames=T,
         color = brewer.pal(11, "PRGn"), 
         main="Heatmap of Genes of Interest",
         fontsize_row = 10,
         cellheight =12, 
         cellwidth =12,
         treeheight_row = 20, 
         treeheight_col = 20, 
         filename = "heatmap_paper_genes.png")
```


### Functional Enrichment Analysis
This script performs Functional Enrichment Analysis using the fgsea package
The results table generated from the Advanced RNA-seq analysis is used as input
The functional data comes from the Molecular Signatures Database (as used for the microarray analysis as well)

```{r gsea, echo=FALSE}
library(fgsea)
library(dplyr)

# We will use the results from the differential expression analysis
results_df <- as.data.frame(results_gsea)
head(results_df)
rownames(results_df)
results_df$ensembl_id <- rownames(results_df)
head(results_df)
```


```{r gsea2, echo=FALSE}
# Reference for code chunk: FGT ICA Example Analysis
#grab the hallmark gene set from bioinf.wehi.edu.au
#generally it is better to cache download results, not download each time, but this file is small!
#You may also wish to check for updates to the annotation.
download.file("https://bioinf.wehi.edu.au/software/MSigDB/mouse_H_v5p2.rdata", destfile = "mouse_H_v5p2.rdata")
load("mouse_H_v5p2.rdata")

#We need to map to EntrezIDs- the annotation uses EntrezID
head(Mm.H, 2)

```


```{r getBM, echo=FALSE}
# Reference for code chunk: FGT ICA Example Analysis
library(biomaRt)
mart <- useDataset("mmusculus_gene_ensembl", mart=useMart("ensembl"))
#listAttributes(mart)
ens2entrez <- getBM(attributes=c("ensembl_gene_id", "entrezgene_id"), mart=mart)
head(ens2entrez)
```


```{r join_id, echo=FALSE}
# Reference for code chunk: Modified from FGT ICA Example Analysis
#Map gene names to entrezgene_id
#if NA remove (we only need to consider mappable genes)

results2 <- inner_join(results_df, ens2entrez, by=join_by("ensembl_id"=="ensembl_gene_id"))

head(results2,n=40)
```


```{r sanity, echo=FALSE}
# Reference for code chunk: Modified from FGT ICA Example Analysis
#sanity check one or two of the results eg

# 5	914.136325	0.52866957	0.1782715	2.96553109	0.0030216080	2.96553109	0.0030216080	0.021383688	ENSMUSG00000051285	319263

ens2entrez[which(ens2entrez$ensembl_gene_id  == "ENSMUSG00000051285"),]

#remove any that are NA
results2 <-results2[is.na(results2$entrezgene_id)==FALSE,]
dim(results2)

head(results2)
```


```{r rank, echo=FALSE}
# Reference for code chunk: Modified from FGT ICA Example Analysis
#This analysis takes a ranked list of numbers with entrez IDs as names
#We need to choose how to order the hits
#Here I order by stat value- we could use bidirectional ordering using FC or some combination of FC and FDR
results2 <-results2[order(results2$stat,decreasing=FALSE),]

results2 <-results2[,c("padj","entrezgene_id")]

#remove non-unique row names- we don't mind deleting these in this case
#it is wise to check the genes that are duplicated to see if they should be excluded
results2 <-results2[!duplicated(results2$entrezgene_id),]


rownames(results2)<-results2[,"entrezgene_id"]
results2["entrezgene_id"] <-NULL


#it might want  a named vector
rn <-rownames(results2)
colnames(results2) <-NULL
results2 <-results2[,1]
names(results2)<-rn

head(results2)

#time to perform the enrichment analysis
results2  <- results2[!(is.na(names(results2)))]
results2  <- results2[!(is.na(results2))]
results2 <-log2(results2)

fgseaRes <- fgsea(Mm.H, results2, minSize=25, maxSize = 500)

fgseaRes <-fgseaRes[order(fgseaRes$NES,decreasing=FALSE),]
head(fgseaRes)

plotEnrichment(Mm.H[["HALLMARK_INFLAMMATORY_RESPONSE"]],results2)

# Save plot
png("fgsea_plot.png", res = 300, width = 2000, height = 1200)
plotEnrichment(Mm.H[["HALLMARK_INFLAMMATORY_RESPONSE"]],results2)
dev.off()
  
```

### GO analysis
Reference for Gene Ontology Analysis: https://www.youtube.com/watch?v=JPwdqdo_tRg
```{r install_packages, echo=FALSE}

if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install("clusterProfiler")

if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install("AnnotationDbi")

if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install("org.Mm.eg.db")

```


```{r load_libraries, echo=FALSE}

library(clusterProfiler)
library(AnnotationDbi)
library(org.Mm.eg.db)

```


```{r select_rows, echo=FALSE}
head(results_df)

head(as.data.frame(result_celltype_selected))

selected <- rownames(result_celltype_selected)
selected
```

```{r go, echo=FALSE}

GO_results <- enrichGO(selected, "org.Mm.eg.db", keyType = "ENSEMBL", ont = "BP")

```

```{r}
# Reference: https://yulab-smu.top/biomedical-knowledge-mining-book/enrichplot.html

as.data.frame(GO_results)

GO_fig <- barplot(GO_results, showCategory = 15, x="GeneRatio", title="Gene Ontology: dLAM vs Mono_M\u03A6")
plot(GO_fig)

png("go_barplot.png", res=300, width=2000, height=3000)
plot(GO_fig)
dev.off()

```



## Session Information
```{r session_info, echo=FALSE}

sessionInfo()

```


