---
title: "RNS-seq Analysis for GSE267325"
author: "B282025"
date: "25/03/2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Analysis of GSE267325 RNA-seq data using R 

```{r init, echo=FALSE}
#Load the required libraries & load the files for the tutorial
library(Rsubread)
library(edgeR)
library(limma)
library(gplots)
library(DESeq2)
library(affy)
library(QuasR)
library(pheatmap)
library(EnhancedVolcano)
load("mytable_feaures")

#We now have files for the tutorial and also data loaded into the workspace
ls()
dir()
```

Next we will use R to generate the targets file.
```{r input_processing, echo=FALSE}
# Some further processing to access count data and column names.
# Reference: FGT ICA Example Analysis
table_rnaseq <- (mytable_feaures)$counts # this is the expression data
colnames_rnaseq <- colnames(table_rnaseq)
print(colnames_rnaseq)

# Generate an 'adf data frame' for the annotation targets file.  
# Reference: FGT ICA Example Analysis
adf <-data.frame(Colnames = colnames_rnaseq)
library(stringr)
locc <- str_locate_all(pattern="Aligned.sortedByCoord.out.bam",adf$Colnames)
adf$GEOAccession <- substr(adf$Colnames,1,locc[[1]][,"start"]-1)

# Sample Annotation from the GEO page for this study: 
# https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE267325

# GSM8263915  dLAM1-CSF
# GSM8263916  dLAM2-CSF
# GSM8263917  dLAM3-CSF
# GSM8263918  Mono-MΦ1-CSF
# GSM8263919  Mono-MΦ2-CSF
# GSM8263920  Mono-MΦ3-CSF

#My samples are in the same order as the annotation, so I will add the short names to the annotation table
# Reference: https://stackoverflow.com/questions/5293715/how-to-use-greek-symbols-in-ggplot2
adf$ShortNames <- c("dLAM1","dLAM2","dLAM3","Mono_M\u03A61","Mono_M\u03A62","Mono_M\u03A63")

# Create a factor for the macrophage type
cell_type <- c("dLAM", "dLAM", "dLAM", "Mono_M\u03A6", "Mono_M\u03A6", "Mono_M\u03A6")
adf$CellType <- cell_type

# Add more annotation data to the adf data frame...
# BioSample
adf$BioSample <- c("SAMN41382413", "SAMN41382412", "SAMN41382411", "SAMN41382410", "SAMN41382409", "SAMN41382408")

# Experiment
adf$Experiment <- c("SRX24727943","	SRX24727944", "SRX24727945", "SRX24727946", "SRX24727947", "SRX24727948")

#Add colours to the annotation table
# Reference: FGT ICA Example Analysis
# Reference: https://stackoverflow.com/questions/42288075/how-to-autogenerate-color-vector-from-a-numeric-vector
z <- factor(adf$CellType)
adf$Colours <-rainbow(length(levels(z)))[z]

# Check the column names of the annotation table (adf)

# Confirm colnames of the study match the rownames of the annotation.
# Reference: FCT ICA Example Analysis
rownames(adf) <-adf$Colnames
if(identical(colnames_rnaseq,rownames(adf))==FALSE){
  print("Names do not match")
}else{
  print("Names match.. OK to proceed")
}

# Switch to shortnames to help with the figures
# Reference: FGT ICA Example Analysis 
#We can only swap this way because column order to consistent with adf
rownames(adf) <-adf$ShortNames
colnames(table_rnaseq) <-adf$ShortNames
colnames_rnaseq <-colnames(table_rnaseq)

# Save the annotated data frame as a targets file
# Reference: Code adapted from Using R for Data Science Lecture 2
write.csv(adf, file="targets_GSE267325.csv", row.names=TRUE)

```
Once we have loaded the data and generated the annotation (saved as targets file) we are ready to run some further analysis.

## Loading the Data into DESeq2
DESeq2 will used for the statistical analysis and the normalisation of the data. 
```{r loading_deseq, echo=FALSE}
# Reference for code chunk: FGT ICA Example Analysis

# Create object for DESeq2
dds_rnaseq <- DESeqDataSetFromMatrix(countData = table_rnaseq, colData = adf, design = ~ CellType)

# Check the object dimensions
dim(dds_rnaseq)
head(rownames(dds_rnaseq))
colnames(dds_rnaseq)

#Check the object type
class(dds_rnaseq)
typeof(dds_rnaseq)

#Check what this object contains
slotNames(dds_rnaseq)
colnames(colData(dds_rnaseq))
colnames(assay(dds_rnaseq))

```

#Quality Filtering Genes

Apply filters to remove genes that do not pass quality filters. In this case, we filter for genes with counts greater than ten.
```{r apply_filters, echo=FALSE}
# First filter based upon counts
# Reference for code chunk: FGT ICA Example Analysis
keep <- rowSums(counts(dds_rnaseq)) > 10
dds_rnaseq <- dds_rnaseq[keep,]
nrow(dds_rnaseq)
```

## Normalisation of Data

```{r normalisation, echo=TRUE}
# Reference for code chunk: FGT ICA Example Analysis
# Size factor estimation for normalisation- these are used to scale by library size when data is exported.
dds_rnaseq <- estimateSizeFactors(dds_rnaseq)
sizeFactors(dds_rnaseq)

```

## DESeq2 also provides methods that outputs data with various normalisations.
Here we generate multiple different normalisations and compare between thenm.

```{r plot_normalised, echo=FALSE}
# Reference for code chunk: FGT ICA Example Analysis
#Normalised log counts
counts_rnaseq <-log2(counts(dds_rnaseq, normalized=TRUE)+1)

#fpm is similar to cpm and discussed here https://rdrr.io/bioc/DESeq2/man/fpm.html.  Here the data is log transformed
fpm_rnaseq <-log(fpm(dds_rnaseq)+1) #this is also normalised by the sizeFactors

#rlog ad vsd are discussed here https://genomebiology.biomedcentral.com/articles/10.1186/s13059-014-0550-8
vsd_rnaseq <- vst(dds_rnaseq, blind = T)
#head(vsd_rnaseq, 3)

rld_rnaseq <- rlog(dds_rnaseq, blind = T)
#head(rld_rnaseq, 3)
```
We will use rlog for the rest of this tutorial, but here we compare the different normalized data.

##Plots with Normalised Data.

```{r build_plots, echo=FALSE}
# Reference for code chunk: FGT ICA Example Analysis
#Boxplot for non-normalized data
counts_un <-counts(dds_rnaseq, normalized=FALSE)+1
boxplot(log2(counts_un), main="Non-Normalized Expression Data", ylab="log2 Expression", col=adf$Colours, las = 2)

counts_un_rlog<-rlog(counts_un)
boxplot(counts_un_rlog, main="Non-Normalized Expression Data", ylab="rlog Expression", col=adf$Colours, las = 2)

# Boxplots for normalized data
na.rm=T
boxplot(counts_rnaseq, main="Normalized Expression Data", ylab="log2 Expression", col=adf$Colours, las = 2)
boxplot(fpm_rnaseq, main="FPM Normalized Expression Data", ylab="log Expression", col=adf$Colours, las = 2)
boxplot(assay(vsd_rnaseq), main=" VST Normalized Expression Data", ylab="log Expression", col=adf$Colours, las = 2)
boxplot(assay(rld_rnaseq), main="rlog Normalized Expression Data", ylab="rlog Expression", col=adf$Colours, las = 2)

```

The non-normalized plots do not present any concerns. The samples and replicates groups all have similar means and interquartile ranges.

Comparing the different normalization techniques, rlog works the best. The rlog normalized data has the most consistent mean and interquartile range spread across samples. For the rest of the pipeline, we will use rlog.

```{r best_boxplot echo=FALSE}
# Save the best plot as png for the report
png("normalized_boxplot.png")
boxplot(assay(rld_rnaseq), main="rlog Normalized Expression Data", ylab="rlog Expression", col=adf$Colours, las = 2, cex.axis = 0.7)
dev.off()
```


#Exploratory Data Plots

First, we will view the MvA plots to compare all sample pairs.
```{r mva_plots, echo=FALSE}
# Reference for code chunk: FGT ICA Example Analysis
# Non-normalized rlog expression
mva.pairs(counts_un_rlog, main="MvA Plot for Non-Normalized rlog Expression Data")

# Normalized rlog expression
mva.pairs(assay(rld_rnaseq), main="MvA Plot for Normalized rlog Expression Data")
```
After normalization, the majority of MvA plots are denser and flatter with points centered around zero, indicating that the normalization worked adequately. 

The authors of the orginal paper used a 2-dimensional PCA.
```{r pca_explore, echo=FALSE}

# Reference for code chunk: FGT ICA Example Analysis
library(ggplot2)

# 2-dimensional PCA plot
# Reference: https://ggplot2.tidyverse.org/reference/geom_text.html#ref-examples
# Reference: https://www.statology.org/remove-legend-ggplot2/ 
# Reference: https://felixfan.github.io/ggplot2-remove-grid-background-margin/ 
ress <-plotPCA(rld_rnaseq, intgroup = c("CellType"))
ress <- ress + 
  geom_text(label = adf$ShortNames,check_overlap=FALSE, size=3, nudge_x=-5, nudge_y=0.5) +
  theme_bw() + 
  labs(title = "PCA Plot", color="Cell Type") + 
  theme(aspect.ratio=1/2) +
  theme(legend.position="none")
plot(ress)

# Save PCA plot as png for report
png("pca_plot.png")
ress <-plotPCA(rld_rnaseq, intgroup = c("CellType"))
ress <- ress + 
  geom_text(label = adf$ShortNames,check_overlap=FALSE, size=3, nudge_x=-5, nudge_y=0.5) +
  theme_bw() + 
  labs(title = "PCA Plot", color="Cell Type") +
  theme(aspect.ratio=1/2) +
  theme(legend.position="none")
plot(ress)
dev.off()

```

```{r 3d_pca, echo=FALSE}
library(scatterplot3d)

# Perform PCA
pca <- prcomp(t(na.omit(assay(rld_rnaseq))), scale=T)

# Plot the PCA results for the first three principle components
s3d<-scatterplot3d(pca$x[,1:3], pch=19, color=adf$Colours, main = "PCA Plot")
s3d.coords <- s3d$xyz.convert(pca$x[,1:3])
text(s3d.coords$x, s3d.coords$y, labels = colnames(rld_rnaseq), pos = 3,offset = 0.5, cex=0.5)
```


## Generate heatmaps with the data

```{r more_heatma-s, echo=FALSE}
library("pheatmap")
library("RColorBrewer")

# Generate a distance matrix
sampleDists <- dist(t(assay(rld_rnaseq)))
#sampleDists

sampleDistMatrix <- as.matrix(sampleDists)
pheatmap(sampleDistMatrix, main="Heatmap of Gene Expression Differences", display_numbers = TRUE)

# Save as png for report
pheatmap(sampleDistMatrix, main="Heatmap of Gene Expression Differences", display_numbers = TRUE, file="sample_differences_heatmap.png")
```
The sample distances represent the differences between gene expression between every pair of samples. We see no difference between one cell type and itself, which makes sense. We see that the cell types cluster together. There is more between group variance than within group variance. The dLAM1 sample and the Mono-Mφ3 sample are the most different. This agrees with what we see on the PCA plot. The cell types cluster together.  

## Differential Gene Expression

Here we perform a two factor analysis using DESeq2.

```{r differential_expression, echo=FALSE}
dds_rnaseq <- DESeq(dds_rnaseq)

#We should apply this here https://support.bioconductor.org/p/77461/#107693
#But there is a bug in this version of DESeq2!!
#dds_rnaseq <- lfcShrink(dds_rnaseq, coef=2)


#This is how you specify a contrast…
# This gives us a table of differentially expressed genes
result_celltype <-results(dds_rnaseq,contrast=c("CellType","dLAM", "Mono_M\u03A6"))


# padj is the p-value with a multiple testing correction
# Plot of differential expression
plotMA(result_celltype, main="DESeq2 Dispersion Plot", ylim=c(-10,10))

# Create a table of the differentially expressed genes with padj values < 0.01
table(result_celltype$padj < 0.01)

# Create a table of the differentially expressed genes with padj values < 0.05
table(result_celltype$padj < 0.05)

# Create a table of significant genes
result_celltype_selected <- subset(result_celltype, padj < 0.05)

# Order the significant genes by fold change
result_celltype_selected <-result_celltype_selected[order(abs(result_celltype_selected$log2FoldChange),decreasing = TRUE), ]
head(result_celltype_selected)
```
We find 3657 genes with significant differential expression at a p-adjusted threshold value of 0.01.

We find 5233 genes with significant differential expression at a p-adjusted threshold value of 0.05.

## Make heatmaps from each of the output tables- selected top 50 or all using fold change and FDR.

```{r differential_expression_heatmaps, echo=FALSE}
top50 <-rownames(result_celltype_selected)[1:50]

# Make a heatmap for the top 50 differentially expressed genes
# Here we are indexing the rlog normalized count data using the rownames of the top 50 significantly differentially expressed genes
pheatmap(assay(rld_rnaseq)[top50,],scale="row",show_rownames=F,main="RLD values, Row Scaled, Top 50")

# Make a heatmap for all of the genes
pheatmap(assay(rld_rnaseq)[rownames(result_celltype_selected),],scale="row",show_rownames=F,main="Treatment Selected, Row Scaled")


```


## Build Gene Annotation. A serialised is generated in case Ensembl is unavailable

```{r build_gene_annotation, echo=TRUE}
#Download Ensembl annotation using BiomaRt and rename the samples
library(biomaRt)
if(file.exists("resultAnnot.RData")){

  load("resultAnnot.RData")
  print("Loaded existing Ensembl annotation...")
}else{
  
  #UK ensembl is being updated  we use a USA mirror, "useast.ensembl.org"
  ensembl_host <-"https://www.ensembl.org"
  head(biomaRt::listMarts(host = ensembl_host), 15)
  head(biomaRt::listAttributes(biomaRt::useDataset(dataset = "mmusculus_gene_ensembl",mart = useMart("ENSEMBL_MART_ENSEMBL",host = ensembl_host))), 40)
  mart <- biomaRt::useDataset(dataset = "mmusculus_gene_ensembl",mart    = useMart("ENSEMBL_MART_ENSEMBL",host    = ensembl_host))
  resultAnnot <- biomaRt::getBM(values=rownames(dds_rnaseq),attributes =     c("ensembl_gene_id","external_gene_name","chromosome_name","start_position","end_position","description","strand"),filters="ensembl_gene_id",mart=mart)
  save(resultAnnot,file="resultAnnot.RData")
}

```

Merge Annotation with Input Data
```{r merge_gene_annotation, echo=FALSE}
#merge with input data
names <- resultAnnot[,1]
resultAnnot <- as.data.frame(resultAnnot)
rownames(resultAnnot) = names
idx<-match(rownames(dds_rnaseq),rownames(resultAnnot))
#make sure annotation is in same order
all(rownames(dds_rnaseq) == rownames(resultAnnot))
grr<-resultAnnot[match(rownames(dds_rnaseq), resultAnnot$ensembl_gene_id),]
all(rownames(dds_rnaseq) == rownames(grr))
resultAnnot <-grr
all(rownames(dds_rnaseq) == rownames(resultAnnot))
#make the nice names
nice_names<-paste(resultAnnot$ensembl_gene_id,resultAnnot$external_gene_name, sep = '_')
resultAnnot$nice_names <-nice_names
head(resultAnnot)
all(rownames(dds_rnaseq) == rownames(resultAnnot))
#check names
rld_rnaseq <- rlog(dds_rnaseq, blind = TRUE)#
idx2 <-match(rownames(result_treatment)[1:50],rownames(dds_rnaseq))
plotme <-(rld_rnaseq)[rownames(result_treatment)[1:50],]
rownames(plotme)<-resultAnnot$nice_names[idx2]
#make heatmap with candidate genes
png(filename="heatmap_candidates.png",width=800, height=1000)
pheatmap(assay(plotme),scale="row",fontsize_row = 10,cellheight =12, cellwidth =12,treeheight_row = 40, treeheight_col = 40)
dev.off()

pheatmap(assay(plotme),scale="row",fontsize_row = 10,cellheight =12, cellwidth =12,treeheight_row = 40, treeheight_col = 40)

```
## Final Results Table
Since we have annotation that matches gene names to candidates we can also annotate the results table from differential
expression

```{r merge_gene_annotation2, echo=FALSE}

#this has the results ordered

#before 
head(result_treatment,50)

idx2 <-match(rownames(result_treatment),rownames(resultAnnot))
rownames(result_treatment)<-resultAnnot$nice_names[idx2]
#after
print(result_treatment[1:50,])

#load into table viewer
#View(as.data.frame(result_treatment[1:50,]))

```

## The source paper showed candidates in a volcano plot- this is to compare

Current row names are not useful here- we need to use genes but then will discard any
without known names. This is OK here because we are exploring the annotation -genes without
a gene name and so are no use for annotating the data.

We can still keep unannoted genes as candidates!

```{r make_volcano, echo=FALSE}

result_treatment$gene<-resultAnnot$external_gene_name[idx2]
has_annot <- result_treatment$gene==""
has_annot <-!has_annot
result_treatment<-result_treatment[has_annot,]
rownames(result_treatment)<-result_treatment$gene

#Now we have gene names so we can plot the remaining genes in a volcano plot

#there were highlighted as interesting in the paper
paper_genes <-c("Zbtb16","Fkbp5","Hif3a","Slc6a3","Sox9","Trem2","Tyrobp","Clec7a","Phox2b")

#just check they're measured in this annotation
if(sum(paper_genes %in% rownames(result_treatment))==length(paper_genes)){
  print("all genes found in data")
}#you'd just include a subset to label if some genes were missing- here they are all present

#The paper found...
#for volvano see 
#https://bioconductor.org/packages/devel/bioc/vignettes/EnhancedVolcano/inst/doc/EnhancedVolcano.html
volcano_test <-EnhancedVolcano(result_treatment,
    lab = rownames(result_treatment),
    x = 'log2FoldChange',
    y = 'padj',
    selectLab = paper_genes,
    xlab = bquote(~Log[2]~ 'fold change'),
    pCutoff = 10e-5,
    FCcutoff = 2.0,
    pointSize = 4.0,
    labSize = 6.0,
    labCol = 'black',
    labFace = 'bold',
    boxedLabels = TRUE,
    colAlpha = 4/5,
    legendPosition = 'right',
    legendLabSize = 10,
    legendIconSize = 4.0,
    drawConnectors = TRUE,
    widthConnectors = 1.0,
    colConnectors = 'black')

plot(volcano_test,xlim=800,ylim=800)
png("volcano_test.png",width = 800, height = 800)
plot(volcano_test)
dev.off()

#and look at the table
#View(as.data.frame(result_treatment[paper_genes,]))

print(as.data.frame(result_treatment[paper_genes,]))


```

## Session Information
```{r session_info, echo=FALSE}

sessionInfo()

```



