---
title: "RNS-seq Analysis for GSE267325"
author: "B282025"
date: "25/03/2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Analysis of RNA-seq data using R 
```{r init, echo=FALSE}
#Load the required libraries & load the files for the tutorial
library(Rsubread)
library(edgeR)
library(limma)
library(gplots)
library(DESeq2)
library(affy)
library(QuasR)
library(pheatmap)
library(EnhancedVolcano)
load("mytable_feaures")

#We now have files for the tutorial and also data loaded into the workspace- use ls() and dir() to view this data
ls()
dir()
```

Once we have loaded the data we are ready to run some further analysis.

```{r input_processing, echo=FALSE}
#Some further processing to access count data and column names.
table_rnaseq <- (mytable_feaures)$counts
colnames_rnaseq <- colnames(table_rnaseq)
print(colnames_rnaseq)

# Generate an 'adf data frame' for the annotation.  
#This can be loaded from file but here I generate in code for a change...
adf <-data.frame(Colnames =colnames_rnaseq)
library(stringr)
locc <- str_locate_all(pattern="Aligned.sortedByCoord.out.bam",adf$Colnames)
adf$ShortGEOName <-substr(adf$Colnames,1,locc[[1]][,"start"]-1)

#From the GEO page for this study-https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE267325

# GSM8263915  dLAM1-CSF
# GSM8263916  dLAM2-CSF
# GSM8263917  dLAM3-CSF
# GSM8263918  Mono-MΦ1-CSF
# GSM8263919  Mono-MΦ2-CSF
# GSM8263920  Mono-MΦ3-CSF


#My samples are in the same order as the annotaion so I can just create some short names and add to annotation
adf$ShortNames <-c("dLAM1","dLAM2","dLAM3","Mono-MΦ1","Mono-MΦ2","Mono-MΦ3")

# Create a factor for the macrophage type
cell_type <-c("dLAM", "dLAM", "dLAM", "Mono-MΦ", "Mono-MΦ", "Mono-MΦ")
adf$CellType <- cell_type

#Add some colours as well
#See https://stackoverflow.com/questions/42288075/how-to-autogenerate-color-vector-from-a-numeric-vector
z <- factor(adf$CellType)
adf$Colours <-rainbow(length(levels(z)))[z]


#make sure the colnames of the study match the rownames of the annotation...
rownames(adf) <-adf$Colnames
if(identical(colnames_rnaseq,rownames(adf))==FALSE){
  print("Names do not match")
}else{
  print("Names match.. OK to proceed")
}

#OK switch to shortnames to help with the figures
#We can only swap this way because column order to consistent with adf
rownames(adf) <-adf$ShortNames
colnames(table_rnaseq) <-adf$ShortNames

colnames_rnaseq <-colnames(table_rnaseq)

```


## Loading the Data into DESeq
DESeq2 will used for the statistical analysis and the normalisation of the data. 
```{r loading_deseq, echo=FALSE}
#Create object for DESeq
#NOTE the design for differential analysis is fed into this object 
dds_rnaseq <- DESeqDataSetFromMatrix(countData = table_rnaseq,colData = adf,design = ~ CellType)
#check the object dimensions
dim(dds_rnaseq)
head(rownames(dds_rnaseq))
colnames(dds_rnaseq)
#check the object type
class(dds_rnaseq)
typeof(dds_rnaseq)

#check what this object contains
slotNames(dds_rnaseq)
colnames(colData(dds_rnaseq))
colnames(assay(dds_rnaseq))

```

Quality Filtering Genes
Apply filters to remove genes that do not pass quality filters.
```{r apply_filters, echo=FALSE}
#first filter based upon counts
keep <- rowSums(counts(dds_rnaseq)) > 1
dds_rnaseq <- dds_rnaseq[keep,]
nrow(dds_rnaseq)
```

## Normalisation of Data

```{r normalisation, echo=TRUE}
#size factor estimation for normalisation- these are used to scale by library size when data is exported.
dds_rnaseq <- estimateSizeFactors(dds_rnaseq)
sizeFactors(dds_rnaseq)

```

## DESeq2 also provides methods that outputs data with various normalisations.
Here we generate multiple different normalisations and compare between then- but for the rest of the tutorial we use rlog normalisation!  Note that the normalised data is not used for differential expression analysis- just for data visualisation.


```{r plot_normalised, echo=FALSE}

#Normalised log counts
counts_rnaseq <-log2(counts(dds_rnaseq, normalized=TRUE))
#fpm is similar to cpm and discussed here https://rdrr.io/bioc/DESeq2/man/fpm.html.  Here the data is log transformed
fpm_rnaseq <-log(fpm(dds_rnaseq))#this is also normalised by the sizeFactors
#rlog ad vsd are discussed here https://genomebiology.biomedcentral.com/articles/10.1186/s13059-014-0550-8
vsd_rnaseq <- vst(dds_rnaseq, blind = T)
head(vsd_rnaseq, 3)
rld_rnaseq <- rlog(dds_rnaseq, blind = T)
head(rld_rnaseq, 3)


```
We will use rlog for the rest of this tutorial, but here we compare the different normalised data

##Plots with Normalised Data.

```{r build_plots, echo=FALSE}

#Boxplots need some labels..
boxplot(counts_rnaseq)
boxplot(fpm_rnaseq)
boxplot(assay(vsd_rnaseq))#etc
boxplot(assay(rld_rnaseq))#etc
#but suppose we look at non-normalised
na.rm=T
counts_un <-counts(dds_rnaseq, normalized=FALSE)+1

boxplot(log2(counts_un))
counts_un_rlog<-rlog(counts_un)





#We can make an mva.pairs plot for any of these data
#mva.pairs(counts_un_rlog,na.rm=TRUE)

```

Exploratory Data Plots
PCA plots similar to those we use for microarrays.


```{r more_explore, echo=FALSE}
#we can run a PCA plot here...
library(scatterplot3d)
library(ggplot2)
# Perform PCA
pca <- prcomp(t(na.omit(assay(rld_rnaseq))), scale=T)
# Plot the PCA results
s3d<-scatterplot3d(pca$x[,1:3], pch=19, color=adf$Colours)
s3d.coords <- s3d$xyz.convert(pca$x[,1:3])
text(s3d.coords$x, s3d.coords$y, labels = colnames(rld_rnaseq),pos = 3,offset = 0.5,cex=0.5)
#or a quick plot in 2d
qplot(pca$x[,1],pca$x[,2],xlab="PCA1", ylab="PCA2",color=adf$Colours)
#We can also output a distance matrix…
sampleDists <- dist(t(assay(rld_rnaseq)))
sampleDists

```
## Generate further heatmaps with the data

```{r more_heatma-s, echo=FALSE}
library("pheatmap")
library("RColorBrewer")
sampleDistMatrix <- as.matrix(sampleDists)
pheatmap(sampleDistMatrix)
#easier PCA plot
ress <-plotPCA(rld_rnaseq, intgroup = c("CellType"))
ress <- ress + geom_label(aes(label = name))
plot(ress)
```


## Differential Gene Expression
Simplified stats

```{r differential_expression, echo=FALSE}
dds_rnaseq <- DESeq(dds_rnaseq)

#We should apply this here https://support.bioconductor.org/p/77461/#107693
#But there is a bug in this version of DESeq2!!
#dds_rnaseq <- lfcShrink(dds_rnaseq, coef=2)


#This is how you specify a contrast…
result_celltype <-results(dds_rnaseq,contrast=c("CellType","dLAM", "Mono-MΦ"))



#make a nice plot of differential expression
plotMA(result_treatment, main="DESeq2 dispersion plot", ylim=c(-2,2))
table(result_treatment$padj < 0.01)
table(result_treatment$padj < 0.05)
result_treatment_selected <- subset(result_treatment, padj < 0.05)
result_treatment <-result_treatment[order(abs(result_treatment$stat),decreasing = TRUE), ]
head(result_treatment)
```

## Make heatmaps from each of the output tables- selected top 50 or all using fold change and FDR.

```{r differential_expression_heatmaps, echo=FALSE}
top70 <-rownames(result_treatment)[1:70]
#make a heatmap for the top 70
pheatmap(assay(rld_rnaseq)[top70,],scale="row",show_rownames=F,main="RLD values, Row Scaled, Top 70")
#all
pheatmap(assay(rld_rnaseq)[rownames(result_treatment),],scale="row",show_rownames=F,main="Treatment Selected, Row Scaled")


```


## Build Gene Annotation. A serialised is generated in case Ensembl is unavailable

```{r build_gene_annotation, echo=TRUE}
#Download Ensembl annotation using BiomaRt and rename the samples
library(biomaRt)
if(file.exists("resultAnnot.RData")){

  load("resultAnnot.RData")
  print("Loaded existing Ensembl annotation...")
}else{
  
  #UK ensembl is being updated  we use a USA mirror, "useast.ensembl.org"
  ensembl_host <-"https://www.ensembl.org"
  head(biomaRt::listMarts(host = ensembl_host), 15)
  head(biomaRt::listAttributes(biomaRt::useDataset(dataset = "mmusculus_gene_ensembl",mart = useMart("ENSEMBL_MART_ENSEMBL",host = ensembl_host))), 40)
  mart <- biomaRt::useDataset(dataset = "mmusculus_gene_ensembl",mart    = useMart("ENSEMBL_MART_ENSEMBL",host    = ensembl_host))
  resultAnnot <- biomaRt::getBM(values=rownames(dds_rnaseq),attributes =     c("ensembl_gene_id","external_gene_name","chromosome_name","start_position","end_position","description","strand"),filters="ensembl_gene_id",mart=mart)
  save(resultAnnot,file="resultAnnot.RData")
}

```

Merge Annotation with Input Data
```{r merge_gene_annotation, echo=FALSE}
#merge with input data
names <- resultAnnot[,1]
resultAnnot <- as.data.frame(resultAnnot)
rownames(resultAnnot) = names
idx<-match(rownames(dds_rnaseq),rownames(resultAnnot))
#make sure annotation is in same order
all(rownames(dds_rnaseq) == rownames(resultAnnot))
grr<-resultAnnot[match(rownames(dds_rnaseq), resultAnnot$ensembl_gene_id),]
all(rownames(dds_rnaseq) == rownames(grr))
resultAnnot <-grr
all(rownames(dds_rnaseq) == rownames(resultAnnot))
#make the nice names
nice_names<-paste(resultAnnot$ensembl_gene_id,resultAnnot$external_gene_name, sep = '_')
resultAnnot$nice_names <-nice_names
head(resultAnnot)
all(rownames(dds_rnaseq) == rownames(resultAnnot))
#check names
rld_rnaseq <- rlog(dds_rnaseq, blind = TRUE)#
idx2 <-match(rownames(result_treatment)[1:50],rownames(dds_rnaseq))
plotme <-(rld_rnaseq)[rownames(result_treatment)[1:50],]
rownames(plotme)<-resultAnnot$nice_names[idx2]
#make heatmap with candidate genes
png(filename="heatmap_candidates.png",width=800, height=1000)
pheatmap(assay(plotme),scale="row",fontsize_row = 10,cellheight =12, cellwidth =12,treeheight_row = 40, treeheight_col = 40)
dev.off()

pheatmap(assay(plotme),scale="row",fontsize_row = 10,cellheight =12, cellwidth =12,treeheight_row = 40, treeheight_col = 40)

```
## Final Results Table
Since we have annotation that matches gene names to candidates we can also annotate the results table from differential
expression

```{r merge_gene_annotation2, echo=FALSE}

#this has the results ordered

#before 
head(result_treatment,50)

idx2 <-match(rownames(result_treatment),rownames(resultAnnot))
rownames(result_treatment)<-resultAnnot$nice_names[idx2]
#after
print(result_treatment[1:50,])

#load into table viewer
#View(as.data.frame(result_treatment[1:50,]))

```

## The source paper showed candidates in a volcano plot- this is to compare

Current row names are not useful here- we need to use genes but then will discard any
without known names. This is OK here because we are exploring the annotation -genes without
a gene name and so are no use for annotating the data.

We can still keep unannoted genes as candidates!

```{r make_volcano, echo=FALSE}

result_treatment$gene<-resultAnnot$external_gene_name[idx2]
has_annot <- result_treatment$gene==""
has_annot <-!has_annot
result_treatment<-result_treatment[has_annot,]
rownames(result_treatment)<-result_treatment$gene

#Now we have gene names so we can plot the remaining genes in a volcano plot

#there were highlighted as interesting in the paper
paper_genes <-c("Zbtb16","Fkbp5","Hif3a","Slc6a3","Sox9","Trem2","Tyrobp","Clec7a","Phox2b")

#just check they're measured in this annotation
if(sum(paper_genes %in% rownames(result_treatment))==length(paper_genes)){
  print("all genes found in data")
}#you'd just include a subset to label if some genes were missing- here they are all present

#The paper found...
#for volvano see 
#https://bioconductor.org/packages/devel/bioc/vignettes/EnhancedVolcano/inst/doc/EnhancedVolcano.html
volcano_test <-EnhancedVolcano(result_treatment,
    lab = rownames(result_treatment),
    x = 'log2FoldChange',
    y = 'padj',
    selectLab = paper_genes,
    xlab = bquote(~Log[2]~ 'fold change'),
    pCutoff = 10e-5,
    FCcutoff = 2.0,
    pointSize = 4.0,
    labSize = 6.0,
    labCol = 'black',
    labFace = 'bold',
    boxedLabels = TRUE,
    colAlpha = 4/5,
    legendPosition = 'right',
    legendLabSize = 10,
    legendIconSize = 4.0,
    drawConnectors = TRUE,
    widthConnectors = 1.0,
    colConnectors = 'black')

plot(volcano_test,xlim=800,ylim=800)
png("volcano_test.png",width = 800, height = 800)
plot(volcano_test)
dev.off()

#and look at the table
#View(as.data.frame(result_treatment[paper_genes,]))

print(as.data.frame(result_treatment[paper_genes,]))


```

## Session Information
```{r session_info, echo=FALSE}

sessionInfo()

```



